\large How resource abundance and resource stochasticity affect organisms' range sizes

### 2025-02-07 writing/manuscript.Rmd
started responding to comments from reviewers

---
title: \large How resource abundance and stochasticity affect organisms' range sizes
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'movement-ecology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character \U0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "\\textbf{Background:} The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. Since organisms adjust to variable environmental conditions, failing to consider the effects of resource (un)predictability can result in an insufficient understanding of an organism's range size, which results in challenges for both ecological theory and applied conservation. \\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes. We then use simulated movement data using continuous-time movement models to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size. Finally, we test the two hypotheses using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado. \\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity. Additionally, the effects of resource stochasticity depend strongly on resource abundance, and organisms' responses to resource abundance and stochasticity can be complex, since the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation. \\textbf{Conclusions:} Accounting for resource abundance and stochasticity is crucial for understanding the movement behavior of free-ranging organisms. Failing to account for resource stochasticity can lead to an incomplete and incorrect understanding of how and why organisms move, particularly during periods of rapid change."
# stringi::stri_count_words(abstract)
```

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met.

While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008]. The relationship between organisms' movement and resource abundance has long been of interest to biologists. In his seminal paper, Burt @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, Southwood @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce. Two years later, Harestad and Bunnel @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption rate (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis. Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings. Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024]. Thus, there remains a need for clear hypotheses of the effects of both resource abundance and stochasticity on organisms’ range sizes.

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. To see this, consider, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance). This individual will adopt different behaviors and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity). In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019]. Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004]. Thus, we can treat the amount of resources $R$ at a given point in time ($t$) and space (location vector $\vec u$) as a random variable, denoted as $R(t, \vec u)$. Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean. We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ to indicate them as functions of time ($t$) and space ($\vec u$). Appendix A defines and expands on the concepts of probability distributions, expected value, variance, and provides examples of them for gamma and beta distributions.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for the growth, survival, and reproduction of most organisms. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. As the abundance of essential resources fluctuates, motile organisms can move to new locations or 'patches' to meet their requirements [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for costs of movement [@taylor_energetics_1982].

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses, which is similar to that presented by @harestad_home_1979. When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$. These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, switches from range residency to nomadism (or vice-versa) will occur over evolutionary timescales rather than over an organism's lifespan (Fig. \@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

Overall, the hypothesis that range size decreases with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because, conceptually, the relationship between range size and $\e(R)$ must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero [see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above]. Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size.

@@

\noindent Assuming resource stochasticity is constant over time and space can be a useful simplification of relatively stable environments or when information on how $\e(R)$ changes is limited and estimating changes in $\var(R)$ is unreasonable. However, such an assumption is likely not realistic, since $\var(R)$ often differ across space and over time. Generally, bounded qualities quantities have correlated means and variances, as in the case of random variables that are strictly positive (e.g., Gamma and Poisson) or fully bounded (e.g., beta). See Appendix A for more information.

@@

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on on the mean (A-C) and variance (D-F) in simulated home-range size with 95\\% Bayesian credible intervals. While the estimated marginal effect of $\\var(R)$ on $\\e(H)$ is sublinear (panel B), the effect of $\\var(R)$ is superlinear for high values of $\\e(R)$ (panel C). The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$. Credible intervals were calculated using 10,000 samples from the posterior distribution while assuming multivariate Gaussian coefficients. Additional details on the model structure are provided in Appendix B.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-mean-and-var.png')
```

@@

\noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move. This paper builds on earlier theoretical work [e.g., @harestad_home_1979; @stephens_optimal_1982; @southwood_habitat_1977] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes. We use quantitative simulations and an empirical case study to support the hypotheses and show that, together, they provide a simple framework for understanding how motile organisms adapt their movement in dynamic environments. Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: $H$ decreases with $\e(R)$ and increases with $\var(R)$. Together, the degree to which $\e(R)$ affects $H$ depends on $\var(R)$, and vice-versa, so organisms' responses to resource dynamics can be complex. The simulated and empirical results suggest qualitatively similar marginal effects of $\e(R)$ and $\var(R)$, but there are differences in the estimated interactive effects. In the simulated data, $\var(R)$ has little effect when $\e(R)$ is low and a strong effect when $\e(R)$ is high, while the opposite is true for the empirical data. This difference is due to two reasons. Firstly, the shape and symmetry of bounded distributions such as Gamma ($R > 0$) and Beta ($0 < R < 1$) distributions depend on both $\e(R)$ and $\var(R)$ (figs. A3, A4), but $\var(R)$ does not affect the shape of a Gamma distribution as much if $\e(R)$ is low (fig. B3). Secondly, and perhaps more interestingly, the simulation approach does not account for real-world adaptations to $\e(R)$ and $\var(R)$ such as selective space use, which we account for in the empirical approach. Below we discuss the strengths and limitations of each approach.

@@

Similarly, if $R$ is often non-measurable. Proxies of $R$, such as NDVI [@pettorelli_normalized_2011], which may introduce complexities. While $R$ and NDVI are correlated for many species [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016]. This complexity can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig. C3) and satellite-level confounding variables ($S$ in fig. C3). Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and NDVI [@jetz_scaling_2004; @prox_framework_2020]. Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators [@duncan_lifehistory_2015], which may, in turn, increase pressures from competition and predation [@charnov_optimal_1976; @brown_ecology_1999]. However, such pressures may result in both an expansion of the range [@jetz_scaling_2004; @prox_framework_2020] or a contraction, since larger ranges can be harder to defend and result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004] and encounter rates [@martinez-garcia_how_2020]. Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@huang_commentary_2021; @fan_global_2016], satellite-level error [@huang_commentary_2021; @tian_evaluating_2015; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources or small-scale resource depletion). However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of remotely-sensed  $R$ [e.g., young grass rather than mature grasslands, see @huang_commentary_2021].

>>>>

---
title: \large How resource abundance and resource stochasticity affect organisms' range sizes
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'movement-ecology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
#   bookdown::word_document2:
#     pandoc_args:
# # https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
#       - '--lua-filter=scholarly-metadata.lua'
#       - '--lua-filter=author-info-blocks.lua'
#     toc: false
#     number_sections: false
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character \U0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
abstract <- "\\textbf{Background:} From megafauna to amoebas, the amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to spatiotemporal changes in mean resources, while responses to unpredictable changes in resources (i.e., variance in resources or resource stochasticity) are still largely unknown. Since organisms adjust to variable environmental conditions, failing to consider the effects of resource unpredictability can result in an insufficient understanding of an organism's range size. \\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on organisms' range sizes. We then use simulated movement data to demonstrate how the combined effects of resource abundance and stochasticity interact to shape predictable patterns in range size. Finally, we test the two hypotheses using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado. \\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity, and the effects of resource stochasticity depend strongly on resource abundance. Additionally, the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation. \\textbf{Conclusions:} Accounting for resource abundance and stochasticity is crucial for understanding the movement behavior of free-ranging organisms. Failing to account for resource stochasticity can lead to an incomplete and incorrect understanding of how and why organisms move, particularly during periods of rapid change."
# stringi::stri_count_words(abstract)
```

@@

\noindent The amount of resources an organism is able to access is a strong determinant of its fitness. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met [e.g., soil amoebae *Dictyostelium spp.*: @van_haastert_food_2009; plants: @taub_root_1996; and animals: @harestad_home_1979].

While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that motile species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008]. The relationship between organisms' movement and resource abundance has long been of interest to biologists. In his seminal paper, Burt @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, Southwood @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce. Two years later, Harestad and Bunnel @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption rate (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis. Many researchers have since demonstrated that organisms adapt their range sizes in response to resource abundance, but results are typically reported as independent, novel findings. Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to changes in mean resource abundance, there is little information on how organisms respond to unpredictable changes in resources [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024]. Thus, there remains a need for a clear, unifying hypothesis of the effects of both resource abundance and stochasticity on organisms’ range sizes.

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. To see this, consider, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance). The animal may require a large but constant home range size as it moves between patches in search of food. If, instead, it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low long-term mean and high stochasticity), it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; $ $ see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019]. Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].

This individual will adopt different behaviors and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity). In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches

@@

\noindent Resources (e.g., food, water, shelter, heat) are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004]. Thus, we can treat the amount of resources $R$ at a given point in time ($t$) and space (location vector $\vec u$) as a random variable, denoted as $R(t, \vec u)$. Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean. We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ to indicate them as functions of time ($t$) and space ($\vec u$). Appendix A defines and expands on the concepts of probability distributions, expected value, variance, and provides examples of them for Gamma and Beta distributions.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for the growth, survival, and reproduction of most organisms. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. As the abundance of essential resources fluctuates, motile organisms can move to new locations or 'patches' to meet their requirements [@charnov_optimal_1976; @brown_ecology_1999], but movement also increases energetic needs [@taylor_energetics_1982].

When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [see Fig. \@ref(fig:mu-hyp) as well as: @herfindal_prey_2005; @nilsen_can_2005; @relyea_home_2000]. Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$. These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale switches from range residency to nomadism (or vice-versa) are more likely to occur over evolutionary timescales rather than over an organism's lifespan (Fig. \@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

Overall, the hypothesis that range size decreases with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @rizzuto_forage_2021; @relyea_home_2000; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because, conceptually, the relationship between range size and $\e(R)$ must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero [see @herfindal_prey_2005; @nilsen_can_2005; @lucherini_habitat_1996; @watson_ferruginous_2020; @simcharoen_female_2014, and contrast them to the earlier references that assume a linear relationship between $H$ and $R$]. Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size.

@@

\noindent Assuming resource stochasticity is constant over time and space can be a useful simplification of relatively stable environments or when information on how $\e(R)$ changes is limited and estimating changes in $\var(R)$ is unreasonable. However, such an assumption is likely not realistic, since $\var(R)$ often differ across space and over time. Generally, bounded quantities have correlated means and variances, as in the case of random variables that are strictly positive (e.g., Gamma and Poisson) or fully bounded (e.g., Beta). For example, prey abundance in a given area over time may approximately follow a Poisson distribution, which implies that the mean and variance will be approximately equal. When prey are scarce (e.g., a mean of 10), the variance will also be low, and when prey are abundant (e.g., a mean of 300) the variance will also be high. This occurs because the behavior, fitness, and predator-prey dynamics of ~300 prey are more stochastic than those of 10 prey [@campillo_effect_2012]. Similarly, in the case of fully bounded random variables, the variance is generally lowest when the mean is near either boundary. For example, successful predation events are predictably scarce if the probability of capture is near 0, predictably common if the probability is near 1, and most stochastic if the probability is near 0.5 [i.e., as far as possible from both 0 and 1; see @lee_effects_2011]. See Appendix A for more information.

@@

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on on the mean (A-C) and variance (D-F) in simulated home-range size with 95\\% Bayesian credible intervals. While the estimated marginal effect of $\\var(R)$ on $\\e(H)$ is sublinear (panel B), the effect of $\\var(R)$ is superlinear for high values of $\\e(R)$ (panel C). The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals}$). Credible intervals were calculated using 10,000 samples from the posterior distribution while assuming multivariate Gaussian coefficients. Additional details on the model structure are provided in Appendix B.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-mean-and-var.png')
```

@@

\noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move. This paper builds on earlier theoretical work [e.g., @southwood_habitat_1977; @harestad_home_1979; @stephens_optimal_1982] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes. We use quantitative simulations and an empirical case study to support the hypotheses and show that, together, they provide a simple framework for understanding how motile organisms adapt their movement in dynamic environments. Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: $H$ decreases with $\e(R)$ and increases with $\var(R)$. Together, the degree to which $\e(R)$ affects $H$ depends on $\var(R)$, and vice-versa, so organisms' responses to resource dynamics can be complex. The simulated and empirical results suggest qualitatively similar marginal effects of $\e(R)$ and $\var(R)$, but there are differences in the estimated interactive effects. In the simulated data, $\var(R)$ has little effect when $\e(R)$ is low and a strong effect when $\e(R)$ is high, while the opposite is true for the empirical data. This difference is due to two reasons. Firstly, the shape and symmetry of bounded distributions such as Gamma ($R > 0$) and Beta ($0 < R < 1$) distributions depend on both $\e(R)$ and $\var(R)$ (figs. A3, A4), but $\var(R)$ does not affect the shape of a Gamma distribution as much if $\e(R)$ is low (fig. B3). Secondly, and perhaps more interestingly, the simulation approach does not account for real-world adaptations to $\e(R)$ and $\var(R)$ such as selective space use, which we account for in the empirical approach. Below we discuss the strengths and limitations of each approach.

@@

Similarly, if $R$ is often non-measurable. Proxies of $R$, such as NDVI [@pettorelli_normalized_2011], which may introduce complexities. While $R$ and NDVI are correlated for many species [e.g., @middleton_green-wave_2018; @geremia_migrating_2019; @merkle_spatial_2019; @phillips_evaluating_2008; @seigle-ferrand_systematic_2021; @merkle_large_2016], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016]. This complexity can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig. C3) and satellite-level confounding variables ($S$ in fig. C3). Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and NDVI [@jetz_scaling_2004; @prox_framework_2020]. Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators [@duncan_lifehistory_2015], which may, in turn, increase pressures from competition and predation [@charnov_optimal_1976; @brown_ecology_1999]. However, such pressures may result in both an expansion of the range [@jetz_scaling_2004; @prox_framework_2020] or a contraction, since larger ranges can be harder to defend and result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004] and encounter rates [@martinez-garcia_how_2020]. Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@huang_commentary_2021; @fan_global_2016], satellite-level error [@huang_commentary_2021; @tian_evaluating_2015; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources or small-scale resource depletion). However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of remotely-sensed  $R$ [e.g., young grass rather than mature grasslands, see @huang_commentary_2021].
